Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-06-29 Manuscript/MPESubmission/Revision/response.tex

\item{\textcolor{blue}{\textbf{Specific variation of data per fossil:} More importantly, it doesn't seem like missing data varied among fossil taxa, unless I'm misreading. What we really want to know is what happens when some fossil taxa are fragmentary, as this is the most common empirical situation. Thus, there are morphological characters scored for some living taxa, some fossil taxa, and missing from others. It's unclear if this was simulated. It would be good if the authors could detail the case where, say, all living taxa had morphological data (0 Ml), the matrix was large (0 Mc), and Mf varied *among* fossil taxa, such that some fossils had 100\% coverage, and varied down to, say, 5\%. That's the crucial case here.}}
\end{enumerate}

\section{Specific comments:}
\begin{enumerate}

\item{\textcolor{blue}{\textbf{line 31:} has it really been shown that the majority of macroevolutionary studies focuses on extant species only? The cited references are examples of large scale extant-only phylogenies. To make such a pointed statement, the authors would need to find support from some kind of literature review or meta-analysis comparing the number of paleontological to neontological macroevolutionary studies. Otherwise, rephrase. }}

We revised this statement as follow: "many large-scale macroevolutionary studies focus solely on living species" (line @@@).

@@

\item{\textcolor{blue}{\textbf{line 37 and throughout:} Rephrase so that 'however' does not start the sentence, such as 'To do this, however, we need…'. Sentences should not start with however if the meaning is to contrast with the previous statement (Strunk, W. and White, E.B. 2009. The Elements of Style. NY: Pearson Education. p 48-49).}}

%TG: Adam suggests there is a response to the Strunk and White book saying it's ok to say however. However (see what I did here? ;)) I don't mind changing it.

@@

\item{\textcolor{blue}{\textbf{line 43:} sentence choppy, consider rephrasing or splitting into two sentences.}}

We modified the sentence as follow: "These approaches differ mainly in how they treat the fossil taxa and it's data. One can use the fossil as tips or as nodes in the phylogeny and use it's age only, it's morphology only or it's age and morphology jointly." (line @@@)

@@

\item{\textcolor{blue}{\textbf{line 44:} three alternatives: age only, morph only, or age and morph jointly. }}

See previous change (Specific comment number 3).

\item{\textcolor{blue}{\textbf{line 47:} Simpson 1944 doesn't seem the appropriate reference for maximum parsimony. Consider (Felsenstein 2004; Hennig 1966) }}

We modified the citation as suggested by reviewer 1 (line @@@).

@@

\item{\textcolor{blue}{\textbf{line 53:} omit 'are used as nodes rather than tips in these phylogenies and their'. Fossils aren't really used as nodes in node calibration, the second part of the sentence is more accurate.}}

We modified the sentence to "Because fossil taxa do not usually have available DNA, only fossils' occurrence dates are used to time calibrate phylogenies" (line @@@).

@@

\item{\textcolor{blue}{\textbf{line 66:} omit 'have been successfully applied to empirical data' and just say 'becoming increasingly popular…list of references'. }}

We modified the sentence as follow: "Total Evidence methods is becoming an increasingly popular way of adding fossil taxa to phylogenies (e.g. ...)" (line @@@).

### 2015-06-29 Manuscript/MPESubmission/Revision/TEM_revised_manuscript.tex
minor comments

\newpage
\section{Introduction}
Although most species that have ever lived are now extinct \citep{novacek1992ext,raup1993extinction}, the majority of macroevolutionary studies focus solely on living species (e.g. \citealp{meredithimpacts2011,jetzthe2012}). Ignoring fossil taxa may lead to misinterpretation of macroevolutionary patterns and processes such as the timing of diversification events \citep[e.g.][]{pyrondivergence2011}, relationships among lineages (e.g. \citealp{manosphylogeny2007}) or niche occupancy (e.g. \citealp{pearmanniche2008}). This has led to increasing consensus among evolutionary biologists that fossil taxa should be included in macroevolutionary studies \citep{jacksonwhat2006,quentaldiversity2010,dietlconservation2011,slaterunifying2013,fritzdiversity2013}. However, to do this we need to be able to place living and fossil taxa into the same phylogenies; a task that remains difficult despite recent methodological developments (e.g. \citealp{pyrondivergence2011,ronquista2012,BEASTmaster}).

Up to now, three main approaches have been used to place both living and fossil taxa into phylogenies. These approaches differ mainly in whether they treat fossil taxa as tips or as nodes in the phylogeny, and in which part of the available fossil data is used (i.e. the age of the fossil only or both its age and morphology). Classical cladistic methods use matrices containing morphological data from both living and fossil taxa and treat each taxon as a tip in the phylogeny. Relationships among the taxa are then inferred using optimality criteria such as maximum parsimony \citep{simpson1945}. This approach is commonly used by paleontologists but it ignores the additional molecular data available from living species and does not allow use of probabilistic methods for dealing with phylogenetic uncertainty. Neontologists, on the other hand, more commonly use probabilistic approaches (e.g. Maximum Likelihood or Bayesian methods) based on matrices containing only molecular data from living species. Because fossil taxa do not usually have available DNA, fossils are used as nodes rather than tips in these phylogenies and their occurrence dates are used to time calibrate phylogenies \citep{zuckerkandl1965}. There have been great improvements in the theory and application of these two approaches (e.g. \citealp{bapsta2013,stadlerdating2013,heaththe2013}) as well as much debate about the ``best" approach to use (e.g. \citealp{spencerefficacy2013,wrightbayesian2014}). However neither approach uses all the available data.

A final approach, known as the Total Evidence method, uses matrices containing molecular data from living taxa and morphological data from both living and fossil taxa \citep{eernissetaxonomic1993}. This approach treats every taxon as a tip in the phylogeny, uses the occurrence age of the fossils to time calibrate the phylogeny (known as tip-dating \citealp{ronquista2012}), and allows the use of probabilistic methods for estimating phylogenetic uncertainty \citep{ronquista2012}. Total Evidence methods have been successfully applied to empirical data (e.g. \citealp{pyrondivergence2011,ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014}), and are becoming an increasingly popular way of adding fossil taxa to phylogenies.
However, although the Total Evidence approach seems very promising, there is one big drawback in using this approach: it requires both molecular and morphological data, both of which can be difficult (or impossible) to collect for every living and fossil taxon in the tree.
Morphological data for living taxa are rarely collected when molecular data are available (e.g. \citealp{O'Leary08022013} \textit{vs.} \citealp{meredithimpacts2011}), and for fossil taxa, data can only be collected from features preserved in the fossil record (for example, in vertebrates, the hardest parts of the skeleton are more often preserved than soft parts \citep{sansomfossilization2013}; and molecular data is (nearly) always unavailable. Therefore Total Evidence matrices are likely to contain a lot of missing data that may affect the method's ability to infer correct topologies, branch lengths and support values \citep{salamin2003}.

>>>>

\newpage
\section{Introduction}
Although most species that have ever lived are now extinct \citep{novacek1992ext,raup1993extinction}, the many large-scale macroevolutionary studies focus solely on living species (e.g. \citealp{meredithimpacts2011,jetzthe2012}).
Ignoring fossil taxa may lead to misinterpretation of macroevolutionary patterns and processes such as the timing of diversification events \citep[e.g.][]{pyrondivergence2011}, relationships among lineages (e.g. \citealp{manosphylogeny2007}) or niche occupancy (e.g. \citealp{pearmanniche2008}).
This has led to increasing consensus among evolutionary biologists that fossil taxa should be included in macroevolutionary studies \citep{jacksonwhat2006,quentaldiversity2010,dietlconservation2011,slaterunifying2013,fritzdiversity2013}.
However, to do this we need to be able to place living and fossil taxa into the same phylogenies; a task that remains difficult despite recent methodological developments (e.g. \citealp{pyrondivergence2011,ronquista2012,BEASTmaster}).

Up to now, three main approaches have been used to place both living and fossil taxa into phylogenies.
These approaches differ mainly in how they treat the fossil taxa and it's data. One can use the fossil as tips or as nodes in the phylogeny and use it's age only, it's morphology only or it's age and morphology jointly.
Classical cladistic methods use matrices containing morphological data from both living and fossil taxa and treat each taxon as a tip in the phylogeny. Relationships among the taxa are then inferred using optimality criteria such as maximum parsimony \citep{Hennig1966,felsenstein2004}.
This approach is commonly used by paleontologists but it ignores the additional molecular data available from living species and does not allow use of probabilistic methods for dealing with phylogenetic uncertainty.
Neontologists, on the other hand, more commonly use probabilistic approaches (e.g. Maximum Likelihood or Bayesian methods) based on matrices containing only molecular data from living species.
Because fossil taxa do not usually have available DNA, only fossils' occurrence dates are used to time calibrate phylogenies \citep{zuckerkandl1965}.
There have been great improvements in the theory and application of these two approaches (e.g. \citealp{bapsta2013,stadlerdating2013,heaththe2013}) as well as much debate about the ``best" approach to use (e.g. \citealp{spencerefficacy2013,wrightbayesian2014}).
However neither approach uses all the available data.

A final approach, known as the Total Evidence method, uses matrices containing molecular data from living taxa and morphological data from both living and fossil taxa \citep{eernissetaxonomic1993}.
This approach treats every taxon as a tip in the phylogeny, uses the occurrence age of the fossils to time calibrate the phylogeny (known as tip-dating \citealp{ronquista2012}), and allows the use of probabilistic methods for estimating phylogenetic uncertainty \citep{ronquista2012}.
Total Evidence methods is becoming an increasingly popular way of adding fossil taxa to phylogenies (e.g. \citealp{pyrondivergence2011,ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014}).
However, although the Total Evidence approach seems very promising, there is one big drawback in using this approach: it requires both molecular and morphological data, both of which can be difficult (or impossible) to collect for every living and fossil taxon in the tree.
Morphological data for living taxa are rarely collected when molecular data are available (e.g. \citealp{O'Leary08022013} \textit{vs.} \citealp{meredithimpacts2011}), and for fossil taxa, data can only be collected from features preserved in the fossil record.
For example, in vertebrates, the hardest parts of the skeleton are more often preserved than soft parts \citep{sansomfossilization2013}; and molecular data is (nearly) always unavailable.
Therefore Total Evidence matrices are likely to contain a lot of missing data that may affect the method's ability to infer correct topologies, branch lengths and support values \citep{salamin2003}.
